A general Nullstellensatz for generalized spaces

### 2019-05-05 paper-qcoh.tex
A couple more references

It is not entirely obvious how to correctly formalize
axiom~\eqref{eq:nullstellensatz0} in unadorned (XXX) first-order logic, and in
fact, we do not believe that an entirely faithful formalization is possible.
We will resolve this issue in the body of this note, where we will restate
Theorem~\ref{thm:characterization0} in a more precise form (Theorem~\ref{thm:characterization}).

@@

From the point of view of theory, the appropriate context for the
Nullstellensatz should be determined. We formulate the Nullstellensatz in the
context of geometric theories and their classifying Grothendieck toposes, but
there should be analogues of the Nullstellensatz in other contexts where it
makes sense to speak of ``classifying gadgets'', and these analogues should
shed light on the topos-theoretic version and ideally even explain it from deeper
principles. In particular, preliminary computations indicate that there is a
version of the Nullstellensatz for arithmetic universes, the predicative cousin
of toposes introduced by Joyal~\cite{XXX} and recently an important object of
consideration by Maietti and Vickers~\cite{XXX}.
\medskip

@@

% XXX similarily
Similarly, we cannot construct the subobject of provable sequents by first
constructing an object of raw (arbitrarily-branching) trees and then cutting
down to an object of correctly formed proof trees. Instead, the subobject of
provable sequents can be obtained by intersecting, internally to~$\E$, all
subsets of the set of sequents which are closed under the rules of geometric
logic.\footnote{Equivalently one can appeal to the Knaster--Tarski fixed point
theorem, which is constructively valid in the form we require
here~\cite{bauer-lumsdaine:bourbaki-witt}, to construct the least fixed point
of the operator which, given a set~$M$ of geometric sequents, computes the set of
geometric sequents which can be derived from those in~$M$ in at most one step.
Details on how to carry out such kinds of inductive constructions can be found
in~\cite{blass:induction} and more specifically in
\cite[Chapter~III]{henry:phd}.}

@@

% XXX hyphenation scholium
\begin{proof}[Proof of Theorem~\ref{thm:characterization}]
Any infinitary first-order formula can be simplified to an equivalent geometric
formula -- on the semantic side by repeatedly applying
Scho\-lium~\ref{scholium:nullstellensatz-more-specific}, on the semantic side by
repeatedly applying axiom scheme~\eqref{eq:nullstellensatz}. Hence we are
reduced to the basic fact (Proposition~\ref{prop:basic-truth}) that, for
geometric formulas~$\varphi$, $\Set[\TT] \models
\varphi$ if and only if~$\TT$ proves~$\varphi$.
\end{proof}

@@

\begin{proof}It is an instructive exercise to verify that models of the stated
form are compact. Conversely, let a~$\TT$-model~$M$ be given. Then~$\TT$ is the
% XXX cite Adamek/Rosicky
filtered colimit of all models over~$M$ which are of the stated form. If~$M$ is
compact, the identity on~$M$ factors over such a model. Hence~$M$ is a
retract of such a model and hence itself isomorphic to a model of this form.
\end{proof}

@@

* XXX restrict to finitary first-order formulas in the scholium

>>>>

It is not entirely obvious how to correctly formalize
axiom~\eqref{eq:nullstellensatz0} in unadorned first-order logic, and in
fact, we do not believe that an entirely faithful formalization is possible.
We will resolve this issue in the body of this note, where we will restate
Theorem~\ref{thm:characterization0} in a more precise form (Theorem~\ref{thm:characterization}).

@@

From the point of view of theory, the appropriate context for the
Nullstellensatz should be determined. We formulate the Nullstellensatz in the
context of geometric theories and their classifying Grothendieck toposes, but
there should be analogues of the Nullstellensatz in other contexts where it
makes sense to speak of ``classifying gadgets'', and these analogues should
shed light on the topos-theoretic version and ideally even explain it from deeper
principles. In particular, preliminary computations indicate that there is a
version of the Nullstellensatz for arithmetic universes, the predicative cousin
of toposes introduced by Joyal and recently an important object of
consideration by Maietti and Vickers~\cite{maietti:au,maietti-vickers:induction,vickers:sketches}.
\medskip

@@

Similarly, we cannot construct the subobject of provable sequents by first
constructing an object of raw (arbitrarily-branching) trees and then cutting
down to an object of correctly formed proof trees. Instead, the subobject of
provable sequents can be obtained by intersecting, internally to~$\E$, all
subsets of the set of sequents which are closed under the rules of geometric
logic.\footnote{Equivalently one can appeal to the Knaster--Tarski fixed point
theorem, which is constructively valid in the form we require
here~\cite{bauer-lumsdaine:bourbaki-witt}, to construct the least fixed point
of the operator which, given a set~$M$ of geometric sequents, computes the set of
geometric sequents which can be derived from those in~$M$ in at most one step.
Details on how to carry out such kinds of inductive constructions can be found
in~\cite{blass:induction} and more specifically in
\cite[Chapter~III]{henry:phd}.}

@@

\begin{proof}[Proof of Theorem~\ref{thm:characterization}]
Any infinitary first-order formula can be simplified to an equivalent geometric
formula -- on the semantic side by repeatedly applying
Scho\-li\-um~\ref{scholium:nullstellensatz-more-specific}, on the semantic side by
repeatedly applying axiom scheme~\eqref{eq:nullstellensatz}. Hence we are
reduced to the basic fact (Proposition~\ref{prop:basic-truth}) that, for
geometric formulas~$\varphi$, $\Set[\TT] \models
\varphi$ if and only if~$\TT$ proves~$\varphi$.
\end{proof}

@@

\begin{proof}It is an instructive exercise to verify that models of the stated
form are compact (in a slightly different context, this is done
in~\cite[Theorem~3.12]{adamek-rosicky:presentable}). Conversely, let
a~$\TT$-model~$M$ be given. Then~$\TT$ is the
filtered colimit of all models over~$M$ which are of the stated form. If~$M$ is
compact, the identity on~$M$ factors over such a model. Hence~$M$ is a
retract of such a model and hence itself isomorphic to a model of this form.
\end{proof}

@@
